---
title: "Simple Time Series"
author: "U"
date: "2026-05-15"
output: pdf_document
---


We're going to start out with a simple example, to try and understand basic concepts of time series analysis.

Let's say we want to train the simplest model to forecast inflation at time t. We work with the following model

$$x_t = \phi x_{t-1} + \epsilon_t$$

Where:

- $x_t$ is the inflation at month t
- $\phi$ is a constant parameter
- $\epsilon$ is random noise 

By the chat's recommendation, we choose $x_t := log(CPI_t) - log(CPI_{t-1})$

Now in order to obtain in




Code:

Libraries
```{r}
library("readxl")
library("dplyr")
library("tidyr")
```

Data
```{r}
data <- read_excel("hist_cpi-u.xlsx", skip = 3)
data <- data[-1, ] # remove first row
data <- data[, -1] # remove first col
data <- data %>% 
  slice(1:(n()-3)) %>% # remove last 3 lines
  mutate(across(everything(),as.numeric)) # make sure all columns are numeric
```

Data explore block
```{r}
monthly_cpi
```

```{r}
monthly_cpi <- pivot_longer( # a single vector of CPI by month
  data, 
  cols = Jan.:Dec.,
  names_to = "Month",
  values_to = "CPI"
)

monthly_cpi$inflation <- c(NA, diff(log(monthly_cpi$CPI))) # log(CPI_t) - log(CPI_{t-1})
monthly_cpi$yoY <- c(rep(NA, 12), diff(log(monthly_cpi$CPI), lag = 12)) # log(CPI_t) - log(CPI_{t-12})
```

Now the inflation column contains values of the form $log(CPI_t) - log(CPI_{t-1})$ for each month t.
Some basic visuals:
```{r}
plot(monthly_cpi$inflation, type = "l", main = "Monthly Inflation (log diff)", ylab = "x_t", xlab = "time")
```

Now we estimate $\phi$, the parameter of the model:
```{r}
x <- monthly_cpi$inflation
x_lag <- x[-length(x)] # x_{t-1}
x_t <- x[-1] # x_t

model <- lm(x_t ~ x_lag)
coef(model)
```

We get very small intercept. This suggests $x_t \approx \epsilon_t$. specifically, we can't tell much about $x_t$ when we try to model it using only $x_{t-1}$


# Second try: AR(12)

We now try to model $x_t$ using $x_{t-12}$ i.e the value from a year before

```{r}
x_lag12 <- x[1:(length(x)-12)]
x_t <- x[13:length(x)]

model12 <- lm(x_t ~ x_lag12 -1)
summary(model12)
```

# Third try: Year over Year inflation

Define : $\pi_{t}^{12} := log(CPI_t) - log(CPI_{t-12})$
Our model: $\pi_{t}^{12} = \Phi \pi_{t-1}^{12} + \epsilon_t$

```{r}
pi <- monthly_cpi$yoY
pi_lag <- pi[-length(pi)]
pi_t <- pi[-1]

model <- lm(pi_t ~ pi_lag)
summary(model)
```

We get a very stable model, as $\Phi = 0.986\approx 1$, and $\hat{\sigma} = 0.007674$.
Let's try to run a simulation.

# Simulate the future

We use our model to create 10000 possible futures.
To do so, we'll use $\pi_{dec 2025}$ to generate 12 sequential values for the year.
From these simulations, we'll want to estimate the probability $P(max_{t \in 2026} pi_t > K)$ when K is the threshold

```{r}
create_path <- function(phi, sigma, pi_0) {
  
  path <- numeric(12)
  path[1] <- pi_0
  
  for (t in 2:12) {
    eps <- rnorm(1, 0, sigma)
    path[t] <- phi * path[t-1] + eps
  }
  
  return(path)
}

create_paths <- function(n, model, pi_start) {
  phi <- coef(model)[2]
  sigma <- summary(model)$sigma
  sim_max <- numeric(n)
  
  for (j in 1:n) {
    path <- create_path(phi, sigma, pi_start)
    sim_max[j] <- max(path)
  }
  
  return(sim_max)
}

```

Visuals
```{r}
hist(sim_end, breaks = 50, main = "Simulated Dec 2026 Inflation")
mean(sim_end)
sd(sim_end)
quantile(sim_end, c(0.05, 0.5, 0.95))

```

Now we just compare with Polymarket thresholds, to see market over/under pricing:
```{r}
Ks <- c(0.04, 0.045, 0.05, 0.06, 0.08, 0.10)

probs <- sapply(Ks, function(K) mean(sim_max > K))

data.frame(Ks, probs)
```

Trying only 1990-2024 data:
```{r}
# Keep only 1990–2024
recent_cpi <- monthly_cpi[monthly_cpi$Year >= 1990 & monthly_cpi$Year <= 2024, ]

# Recompute YoY inflation (in case NA at start)
recent_cpi$yoY <- c(rep(NA, 12), diff(log(recent_cpi$CPI), lag = 12))

# Drop initial NAs
pi_recent <- recent_cpi$yoY[-(1:12)]

# Build AR(1) model
pi_lag <- pi_recent[-length(pi_recent)]
pi_t <- pi_recent[-1]

model_recent <- lm(pi_t ~ pi_lag)
summary(model_recent)
```

simulation:
```{r}
sim_max <- create_paths(10000, model_recent, tail(pi_recent, 1))
```

results:
```{r}
hist(sim_max, breaks = 50, main = "Simulated Dec 2026 Inflation, recent model")
mean(sim_max)
sd(sim_max)
quantile(sim_max, c(0.05, 0.5, 0.95))
```

```{r}
Ks <- c(0.04, 0.045, 0.05, 0.06, 0.08, 0.10)

probs <- sapply(Ks, function(K) mean(sim_max > K))

data.frame(Ks, probs)
```

# two regime

The basic model only handles simple gaussian noise. We wish to model more extreme events with higher spikes and variance in the noise.
Thus we use a SETAR model, which means we split the series into two regimes, based on a threshold on past inflation:

$$
\pi_t =
\begin{cases}
\phi_1 \pi_{t-1} + \epsilon_{t}^{1}, \space \pi_{t-1} < r \text{ (normal regime)} \\
\phi_2 \pi_{t-1} + \epsilon_{t}^{2}, \space \pi_{t-1} \ge r \text{ (high inflation regime)}
\end{cases}
$$
r is the threshold, epsilon(i) ~ N(0, sigma_i).
to pick r, we should pick a value which represents a high inflation regime.
We arbitrary pick the 90'th quantile from our pi_recent vector
```{r}
r <- quantile(pi_recent, 0.7, na.rm=TRUE)
```

Train model for each regime:
```{r}
# Normal regime
pi_normal <- pi_recent[pi_recent <= r]
pi_normal_lag <- pi_normal[-length(pi_normal)]
pi_normal_t <- pi_normal[-1]

# High-inflation regime
pi_high <- pi_recent[pi_recent > r]
pi_high_lag <- pi_high[-length(pi_high)]
pi_high_t <- pi_high[-1]
```

Fit separate AR(1) models:
```{r}
model_normal <- lm(pi_normal_t ~ pi_normal_lag)
model_high <- lm(pi_high_t ~ pi_high_lag)

summary(model_normal)
summary(model_high)
```



Simulation function:
```{r}
simulate_two_regime <- function(n, pi_0, phi1, sigma1, phi2, sigma2, r) {
  path <- numeric(n)
  path[1] <- pi_0
  
  for (t in 2:n) {
    if (path[t-1] <= r) {
      path[t] <- phi1 * path[t-1] + rnorm(1, 0, sigma1)
    } else {
      path[t] <- phi2 * path[t-1] + rnorm(1, 0, sigma2)
    }
  }
  return(path)
}


```

Simulation
```{r}
# Number of simulations
n_sim <- 10000
# Number of months to simulate (2026)
n_months <- 12
# Starting point: last known YoY inflation
pi_0 <- tail(pi_recent, 1)
# Threshold r (e.g., 90th percentile)
r <- quantile(pi_recent, 0.9, na.rm=TRUE)

# Parameters from the separate AR(1) fits
phi1 <- coef(model_normal)[2]; sigma1 <- summary(model_normal)$sigma
phi2 <- coef(model_high)[2]; sigma2 <- summary(model_high)$sigma

sim_max <- numeric(n_sim)  # store maximum inflation for each simulated path

for (j in 1:n_sim) {
  path <- simulate_two_regime(n_months, pi_0, phi1, sigma1, phi2, sigma2, r)
  sim_max[j] <- max(path)  # maximum YoY inflation in 2026
}

# Polymarket thresholds (in fraction, e.g., 0.04 = 4%)
Ks <- c(0.04, 0.045, 0.05, 0.06, 0.08, 0.10)

probs <- sapply(Ks, function(K) mean(sim_max > K))
data.frame(Threshold = Ks, Probability = probs)
```

entire process for all data for regime models:
```{r}
# YoY inflation
monthly_cpi$yoY <- c(rep(NA,12), diff(log(monthly_cpi$CPI), lag=12))
pi_all <- monthly_cpi$yoY[-(1:12)]

# Threshold r for high inflation (90th percentile)
r <- quantile(pi_all, 0.9, na.rm=TRUE)

# Split into regimes
pi_normal <- pi_all[pi_all <= r]
pi_high   <- pi_all[pi_all > r]

# Fit AR(1) for each regime
phi1 <- coef(lm(pi_normal[-1] ~ pi_normal[-length(pi_normal)]))[2]
sigma1 <- summary(lm(pi_normal[-1] ~ pi_normal[-length(pi_normal)]))$sigma

phi2 <- coef(lm(pi_high[-1] ~ pi_high[-length(pi_high)]))[2]
sigma2 <- summary(lm(pi_high[-1] ~ pi_high[-length(pi_high)]))$sigma

# Two-regime simulation function
simulate_two_regime <- function(n, pi0, phi1, sigma1, phi2, sigma2, r, df=5){
  path <- numeric(n); path[1] <- pi0
  for(t in 2:n){
    if(path[t-1] <= r) path[t] <- phi1*path[t-1] + rt(1, df=df)*sigma1
    else path[t] <- phi2*path[t-1] + rt(1, df=df)*sigma2
  }
  path
}

# Monte Carlo 10k paths
n_sim <- 10000; n_months <- 12; pi0 <- tail(pi_all,1)
sim_max <- replicate(n_sim, max(simulate_two_regime(n_months, pi0, phi1,sigma1,phi2,sigma2,r)))

# Probabilities for Polymarket thresholds
Ks <- c(0.04,0.045,0.05,0.06,0.08,0.10)
data.frame(Ks, probs = sapply(Ks, function(K) mean(sim_max > K)))
```

# stochastic volatility

```{r}
simulate_two_regime_sv <- function(n, pi0, phi1, sigma1, phi2, sigma2, r, tau=0.01) {
  path <- numeric(n)
  path[1] <- pi0
  
  # Initialize regime volatilities
  sigma_t <- if (pi0 <= r) sigma1 else sigma2
  
  for (t in 2:n) {
    # Random walk for volatility
    sigma_t <- sigma_t * exp(rnorm(1, 0, tau))
    
    # Determine regime
    if (path[t-1] <= r) {
      path[t] <- phi1*path[t-1] + rnorm(1, 0, sigma_t)
    } else {
      path[t] <- phi2*path[t-1] + rnorm(1, 0, sigma_t)
    }
  }
  path
}

# Example: simulate 12 months for 1 path
pi_start <- tail(pi_all,1)
simulate_two_regime_sv(12, pi_start, phi1, sigma1, phi2, sigma2, r, tau=0.02)
```

10000 paths:
```{r}
n_sim <- 10000
n_months <- 12

sim_max <- replicate(n_sim, max(simulate_two_regime_sv(n_months, pi_start, phi1, sigma1, phi2, sigma2, r, tau=0.05)))

Ks <- c(0.04,0.045,0.05,0.06,0.08,0.10)
data.frame(Ks, probs = sapply(Ks, function(K) mean(sim_max > K)))
```

